### 2021-11-30 article.tex
Add Dep. Scala preprint

We believe that type-level programming is not necessarily incompatible with other programming paradigms and that the current division exists mainly due to a lack of attention from the research community.
Unfortunately, most of the existing research conducted in this domain is not directly applicable to languages with \emph{subtyping}.
Although the combination of subtyping and type-level programming has been studied extensively on the theoretical side, through the means of dependently typed systems~\citep{aspinall1994singleton, zwanenburg1999pure, stone2000deciding, courant2003strong, hutchins2010pure, yang2017unifying}, the practical side remains largely unexplored.

@@

In this section, we formally present \SystemFm, an extension of \SystemFsub~\citep{cardelli1994extension} with pattern matching, opaque classes, and match types.
\Cref{fig:FMSyntaxEvaluation} defines \Fm's syntax and evaluation relation. \Cref{fig:FMTypingRules} defines \Fm's type system, composed of three relations: typing, subtyping, and type disjointness.
We discuss differences to \SystemFsub in the following subsections (\Cref{subsec:classes} and \ref{subsec:matches}).
In \Cref{subsec:type-safety}, we outline a proof of type safety for \SystemFm.
In \Cref{subsec:type-binding-extension}, we present an extension of \SystemFm with support for binding pattern variables in type patterns.

@@

The Scala compiler checks for pattern matching exhaustivity to prevent runtime exceptions caused by missing cases.
Exhaustivity checking uses static knowledge about the class hierarchy (such as the sealed and final annotations) to check that every value in the scrutinee type is covered by the pattern clauses~\citep{fengyun2016generic}.
Non-exhaustive patterns are compiled with an additional "catch-all" case which throws a runtime exception.
\SystemFm uses default cases as a replacement for systematic exhaustivity checks or runtime exceptions.

@@

Unlike our calculus, the Scala implementation also allows match types to be defined recursively.
Recursive match types can cause subtyping checks to loop indefinitely.
Our implementation does not check match types for termination, as any such check would necessarily limit expressiveness or convenience.
Instead, we detect divergence during match type reduction using a fuel mechanism.
The compiler is given an initial amount of fuel, which is consumed one unit at a time on every reduction step.
If the compiler runs out of fuel, the reduction is aborted with a "recursion limit exceeded" error.
The current implementation uses a fixed amount of initial fuel.
Although this seems to be sufficient for most practical purposes, we plan on making it configurable.
This mechanism is completely standard and already used in other programing languages with unbounded recursion at the type level~\citep{eisenberg2014closed, vilhelm2015dependently, eisenberg2016dependent}.

@@

In this section, we present a case study to show how match types can be used to express complex type constraints, which in turn can prevent certain programming errors at compile time.
To this end, we outline the type-level implementation of a library for multidimensional arrays which mimics the NumPy API~\citep{harris2020numpy}.
The goal of our library is to provide a \emph{shape}-safe interface for manipulating $\n$-dimensional arrays (abbreviated ndarrays), where array shapes and indices are checked for errors at compile-time rather than at runtime.
Shape and indexing errors in ndarrays is a widely acknowledged problem~\citep{barham2019stuck, rush2019namedtensor}, and several solutions have already been proposed, notably in the form of libraries that rely on type-level programming~\citep{chen2017typesafe, huang2017hasktorch}.
Our library uses match types to provide a shape-safe NumPy-like interface.

@@

% \subsection{Related Work}
% Shape and indexing errors in ndarrays is a widely acknowledged problem~\citep{barham2019stuck, rush2019namedtensor}, and several solutions have been suggested.
% The Dex language~\citep{maclaurin2019dex} introduces an experimental language with a Hindley--Milner-based type system with types for index sets, but cannot express variable-rank ndarrays and lacks the type-level arithmetic needed for typing reshapes.
% Nexus~\citep{chen2017typesafe} represents variable-rank ndarrays as HLists of dimension labels, and uses Scala's implicit resolution to express type-level transformations on shapes.
% However, it only guards against rank errors, and not ndarray size errors.

@@

Singleton types provide an interesting middle ground between unified and separate syntax and have also been studied in conjunction with dependent types and subtyping~\citep{aspinall1994singleton, stone2000deciding, courant2003strong}.
Singleton types give a mechanism to refer to terms \emph{in} types, usually by means of a set-like syntax.
This mechanism is appealing because it allows type system designers to cherry-pick the term constructs that should be allowed in types.
When multiple constructs are shared between terms and types, singleton types provide a clear economy of concepts.
In our study of match types, a minimal use of singleton types would result in sharing a single constructor between the term and the type languages: the constructor for matches. It is unclear if the benefits in doing so would outweigh the additional complexity.

Dependent Object Types (DOT) are, to this day, the most significant effort in formalizing Scala's type system~\citep{essence2016amin}.
DOT does not directly support any form of type-level computation.
We considered using DOT as a starting point for our work, however, despite the recent effort to simplify DOT's soundness proof~\citep{rapoport2017simple, giarrusso2020soundness}, extending DOT remains too big of a challenge to concisely describe language extensions.

@@

Haskell's type families~\citep{chakravarty2005associated, schrijvers2008open} allow programmers to define type-level functions using pattern matching.
By default, type families are open, which means that a definition can spread across multiple files and compilations.
This flexibility induces a substantial restriction on type family definitions: patterns must not overlap.
One benefit of this restriction is that it prevents any ambiguity in the reduction of type families (patterns are pairwise disjoint), which is required given the distributed nature of definitions.
Open type families are well-suited to be used in conjunction with type classes, since both constructs have open-ended definitions with non-overlapping constraints.

@@

Although the situation is concerning, it might not be as bad as it seems given that soundness is a non-goal of TypeScript's type system~\citep{bierman2014typescript}.
Nevertheless, we believe that the results of this paper are directly applicable to conditional types and could be used to improve TypeScript's type checker.

>>>>

We believe that type-level programming is not necessarily incompatible with other programming paradigms and that the current division exists mainly due to a lack of attention from the research community.
Unfortunately, most of the existing research conducted in this domain is not directly applicable to languages with \emph{subtyping}.
Although the combination of subtyping and type-level programming has been studied extensively on the theoretical side, through the means of dependently typed systems~\citep{aspinall1994subtyping, zwanenburg1999pure, stone2000deciding, courant2003strong, hutchins2010pure, yang2017unifying}, the practical side remains largely unexplored.

@@

In this section, we formally present \SystemFm, an extension of \SystemFsub~\citep{cardelli1994an} with pattern matching, opaque classes, and match types.
\Cref{fig:FMSyntaxEvaluation} defines \Fm's syntax and evaluation relation. \Cref{fig:FMTypingRules} defines \Fm's type system, composed of three relations: typing, subtyping, and type disjointness.
We discuss differences to \SystemFsub in the following subsections (\Cref{subsec:classes} and \ref{subsec:matches}).
In \Cref{subsec:type-safety}, we outline a proof of type safety for \SystemFm.
In \Cref{subsec:type-binding-extension}, we present an extension of \SystemFm with support for binding pattern variables in type patterns.

@@

The Scala compiler checks for pattern matching exhaustivity to prevent runtime exceptions caused by missing cases.
Exhaustivity checking uses static knowledge about the class hierarchy (such as the sealed and final annotations) to check that every value in the scrutinee type is covered by the pattern clauses~\citep{liu2016a}.
Non-exhaustive patterns are compiled with an additional "catch-all" case which throws a runtime exception.
\SystemFm uses default cases as a replacement for systematic exhaustivity checks or runtime exceptions.

@@

Unlike our calculus, the Scala implementation also allows match types to be defined recursively.
Recursive match types can cause subtyping checks to loop indefinitely.
Our implementation does not check match types for termination, as any such check would necessarily limit expressiveness or convenience.
Instead, we detect divergence during match type reduction using a fuel mechanism.
The compiler is given an initial amount of fuel, which is consumed one unit at a time on every reduction step.
If the compiler runs out of fuel, the reduction is aborted with a "recursion limit exceeded" error.
The current implementation uses a fixed amount of initial fuel.
Although this seems to be sufficient for most practical purposes, we plan on making it configurable.
This mechanism is completely standard and already used in other programing languages with unbounded recursion at the type level~\citep{eisenberg2014closed, sjoberg2015a, eisenberg2016dependent}.

@@

In this section, we present a case study to show how match types can be used to express complex type constraints, which in turn can prevent certain programming errors at compile time.
To this end, we outline the type-level implementation of a library for multidimensional arrays which mimics the NumPy API~\citep{harris2020array}.
The goal of our library is to provide a \emph{shape}-safe interface for manipulating $\n$-dimensional arrays (abbreviated ndarrays), where array shapes and indices are checked for errors at compile-time rather than at runtime.
Shape and indexing errors in ndarrays is a widely acknowledged problem~\citep{barham2019machine, rush2019tensor}, and several solutions have already been proposed, notably in the form of libraries that rely on type-level programming~\citep{chen2017typesafe, huang2017hasktorch}.
Our library uses match types to provide a shape-safe NumPy-like interface.

@@

% \subsection{Related Work}
% Shape and indexing errors in ndarrays is a widely acknowledged problem~\citep{barham2019machine, rush2019tensor}, and several solutions have been suggested.
% The Dex language~\citep{maclaurin2019dex} introduces an experimental language with a Hindley--Milner-based type system with types for index sets, but cannot express variable-rank ndarrays and lacks the type-level arithmetic needed for typing reshapes.
% Nexus~\citep{chen2017typesafe} represents variable-rank ndarrays as HLists of dimension labels, and uses Scala's implicit resolution to express type-level transformations on shapes.
% However, it only guards against rank errors, and not ndarray size errors.

@@

Singleton types provide an interesting middle ground between unified and separate syntax and have also been studied in conjunction with dependent types and subtyping~\citep{aspinall1994subtyping, stone2000deciding, courant2003strong}.
Singleton types give a mechanism to refer to terms \emph{in} types, usually by means of a set-like syntax.
This mechanism is appealing because it allows type system designers to cherry-pick the term constructs that should be allowed in types.
When multiple constructs are shared between terms and types, singleton types provide a clear economy of concepts.
In our study of match types, a minimal use of singleton types would result in sharing a single constructor between the term and the type languages: the constructor for matches. It is unclear if the benefits in doing so would outweigh the additional complexity.

Dependent Object Types (DOT) are, to this day, the most significant effort in formalizing Scala's type system~\citep{amin2017type}.
DOT does not directly support any form of type-level computation.
We considered using DOT as a starting point for our work, however, despite the recent effort to simplify DOT's soundness proof~\citep{rapoport2017a, giarrusso2020scala}, extending DOT remains too big of a challenge to concisely describe language extensions.

@@

Haskell's type families~\citep{chakravarty2005associated, schrijvers2008type} allow programmers to define type-level functions using pattern matching.
By default, type families are open, which means that a definition can spread across multiple files and compilations.
This flexibility induces a substantial restriction on type family definitions: patterns must not overlap.
One benefit of this restriction is that it prevents any ambiguity in the reduction of type families (patterns are pairwise disjoint), which is required given the distributed nature of definitions.
Open type families are well-suited to be used in conjunction with type classes, since both constructs have open-ended definitions with non-overlapping constraints.

@@

Although the situation is concerning, it might not be as bad as it seems given that soundness is a non-goal of TypeScript's type system~\citep{bierman2014understanding}.
Nevertheless, we believe that the results of this paper are directly applicable to conditional types and could be used to improve TypeScript's type checker.
